---
title: "eQTL_AF_Harbin"
author: "Francis Leblanc"
date: "21/01/2023"
output:   
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(2023)
```

# lib 

```{r, echo = FALSE, warning = FALSE, results = 'hide', message = FALSE}
library(parallel)
library(Biobase)
library(snpStats)
library(ggplot2)
library(MatrixEQTL)
library(rio)
library(psych)
library(Hmisc)
library(dplyr)
```


Gathering GWAS data overlaping genotype and RNAseq + liftover (not needed - done in CTSN)
Extract variants in 250kb windows around sentinel (not needed - done in CTSN)
see 03a_eQTL_AF_CTSN.rmd

filter on MAF (0.05), SNPs and individuals to create subsets files and run eQTL analysis.

# Create input tables for MatrixEQTL

## First the genotype table is created from plink files with snpStats package (0,1,2 notation for genotypes)

```{r}
#### With grep, get all matching Ids from bim file ###
#### Make sure line endings are ok!
#### if not use > sed 's/\r//g'
# module load plink/1.9b_6.21-x86_64
# grep -f data/external/AF.snp.locs.table data/Harbin/plink/imputed.bim | cut -f 2 > data/Harbin/plink/AF_vars.txt
```

```{r}
# gene expression matrix for sample order
GE <- readRDS("data/Harbin/RNAseq/vst.rds")
colnames(GE) <- gsub("-", ".", colnames(GE))

fam <- "data/Harbin/plink/eQTL_250kb.fam"
bim <- "data/Harbin/plink/eQTL_250kb.bim"
bed <- "data/Harbin/plink/eQTL_250kb.bed"

sample.harbin <- read.plink(bed, bim, fam)
geno.harbin <- data.frame(t(as(sample.harbin$genotypes, "numeric")))
geno.harbin.o <- data.frame(id = row.names(geno.harbin),
                            geno.harbin[, colnames(GE)])

write.table(geno.harbin.o, 
            "data/Harbin/MatrixEQTL/SNP.txt", 
            row.names = FALSE,  
            quote = F , 
            sep = "\t")
```

## 2nd create expression table

```{r}
GE <- data.frame(id = row.names(GE), GE)
write.table(GE,
            "data/Harbin/MatrixEQTL/expr_g.txt", 
            row.names=FALSE, 
            quote = F,
            sep = "\t")
```

## 3rd create location table of gene for cis query
done in 03a_eQTL_AF_CTSN.rmd

## 4th create location table of SNPs for cis query

```{r}
snpsloc.harbin <- data.frame(snp = sample.harbin$map$snp.name,
                      chr = paste0("chr",sample.harbin$map$chromosome),
                      pos = sample.harbin$map$position)
write.table(snpsloc.harbin,
            "data/Harbin/MatrixEQTL/eQTL_250kb_snpsloc.txt",
            row.names = FALSE, 
            quote = F,
            sep = "\t")
```

## 5th Create covariate table with numerical values

#### 7PC as covariates 

```{r}
meta.harbin <- readRDS("data/Harbin/RNAseq/meta_cov.7PCs.rds")

### code M as 0 and F as 1, Ctl as 0 and AF as 1
meta.harbin$sex <- ifelse(meta.harbin$sex == "Female",  1,  0)
meta.harbin$condition <- ifelse(meta.harbin$condition == "AF",  1,  0)
meta.harbin.t <- t(meta.harbin)

Covariates.harbin <- data.frame(id = row.names(meta.harbin.t), 
                                meta.harbin.t)
Covariates.harbin_noAF <- Covariates.harbin[-c(2), ]
write.table(Covariates.harbin_noAF,
            "data/Harbin/MatrixEQTL/Covariates_noAF_7PCs.txt",
            row.names = FALSE, 
            quote = F,
            sep = "\t")
```

# MatrixEQTL in cis analysis
Now we can lunch the eQTL analysis from MatrixEQTL package
Parameters used:

* linear model
* Cis distance : 1MB
* coaviates : Sex 
* output only paires in trans with padj, < 0.01


### Harbin 7 PCs 250kb window

```{r message=F, warning=F}
base.dir = getwd()
useModel = modelLINEAR
SNP_file_name = paste(base.dir, 
                      "data/Harbin/MatrixEQTL/SNP.txt",
                      sep = "")
snps_location_file_name = paste(base.dir,
                                "data/Harbin/MatrixEQTL/eQTL_250kb_snpsloc.txt", 
                                sep = "")
expression_file_name = paste(base.dir, 
                             "data/Harbin/MatrixEQTL/expr_g.txt", 
                             sep = "")
gene_location_file_name = paste(base.dir,
                                "data/external/geneloc.txt", 
                                sep = "");
covariates_file_name = paste(base.dir,
                             "data/Harbin/MatrixEQTL/Covariates_noAF_7PCs.txt",
                             sep = "")

# Output file name
output_file_name_cis = tempfile();
output_file_name_tra = tempfile();

# Only associations significant at this level will be saved
pvOutputThreshold_cis = 1;
pvOutputThreshold_tra = 0;

errorCovariance = numeric() # might need to remove that

# Distance for local gene-SNP pairs
cisDist = 1e6;

## Load genotype data

snps = SlicedData$new()
snps$fileDelimiter = "\t"      # the TAB character
snps$fileOmitCharacters = "NA" # denote missing values;
snps$fileSkipRows = 1     # one row of column labels
snps$fileSkipColumns = 1     # one column of row labels
snps$fileSliceSize = 2000    # read file in pieces of 2,000 rows
snps$LoadFile( SNP_file_name )

## Load gene expression data

gene = SlicedData$new();
gene$fileDelimiter = "\t";      # the TAB character
gene$fileOmitCharacters = "NA"; # denote missing values;
gene$fileSkipRows = 1;          # one row of column labels
gene$fileSkipColumns = 1;       # one column of row labels
gene$fileSliceSize = 2000;      # read file in slices of 2,000 rows
gene$LoadFile(expression_file_name);

## Load covariates

cvrt = SlicedData$new();
cvrt$fileDelimiter = "\t";      # the TAB character
cvrt$fileOmitCharacters = "NA"; # denote missing values;
cvrt$fileSkipRows = 1;          # one row of column labels
cvrt$fileSkipColumns = 1;       # one column of row labels
if(length(covariates_file_name) > 0) {
cvrt$LoadFile(covariates_file_name);
}

## Run the analysis
snpspos = read.table(snps_location_file_name, 
                     header = TRUE, 
                     stringsAsFactors = FALSE);
genepos = read.table(gene_location_file_name, 
                     header = TRUE, 
                     stringsAsFactors = FALSE);

me = Matrix_eQTL_main(
  snps = snps,
  gene = gene,
  cvrt = cvrt,
  output_file_name = output_file_name_tra,
  pvOutputThreshold = pvOutputThreshold_tra,
  useModel = useModel,
  errorCovariance = errorCovariance,
  verbose = TRUE,
  output_file_name.cis = output_file_name_cis,
  pvOutputThreshold.cis = pvOutputThreshold_cis,
  snpspos = snpspos,
  genepos = genepos,
  cisDist = cisDist,
  pvalue.hist = "qqplot",
  min.pv.by.genesnp = FALSE,
  noFDRsaveMemory = FALSE);

unlink(output_file_name_tra);
unlink(output_file_name_cis);
saveRDS(me, "data/Harbin/MatrixEQTL/250kb_eQTLs_7PCs.RDS")
```


## Annotation

```{r}
me <- readRDS("data/Harbin/MatrixEQTL/250kb_eQTLs_7PCs.RDS")
lead.snps <- read.delim("data/Harbin/plink/AF_vars.txt", header = F) 
fdata <- readRDS("data/external/gencode.v32.annotation.genes.rds")
meta.snps <- rio::import("data/external/sentinel.snp.info.csv")

lead.snps$pos <- limma::strsplit2(lead.snps$V1,":")[, 2]
meta.snps$plink_id <- lead.snps$V1[match(meta.snps$CHR_POS, lead.snps$pos)]

cis_eqtls  <- filter(me$cis$eqtls, snps %in% lead.snps$V1)
cis_eqtls$GENEID <- limma::strsplit2(cis_eqtls$gene, "[.]")[,1]
cis_eqtls <- cbind(cis_eqtls,
                   fdata[match(cis_eqtls$GENEID, fdata$ensembl_gene_id),
                         c("gene_name", "seqnames","start","gene_type")])

cis_eqtls$snps_gene <- paste0(cis_eqtls$snps, "_", cis_eqtls$gene_name)
cis_eqtls[, c("rsid","Meta_P")] <- meta.snps[match(cis_eqtls$snps, 
                                                  meta.snps$plink_id),
                                            c("SNPS","P-VALUE")]

saveRDS(cis_eqtls, "results/Harbin_cis_eQTLs.sex.7PCs.RDS")
```


##  qq plot with GGplot (Figure 1B)

```{r}
cis_eqtls <- readRDS("results/Harbin_cis_eQTLs.sex.7PCs.RDS")

df.ci<- data.frame(
    expected = -log10(ppoints(nrow(cis_eqtls))),
    clower   = -log10(qbeta(p = (1 - 0.95) / 2, shape1 = 1:nrow(cis_eqtls), 
                            shape2 = nrow(cis_eqtls):1)),
    cupper   = -log10(qbeta(p = (1 + 0.95) / 2, shape1 = 1:nrow(cis_eqtls), 
                            shape2 = nrow(cis_eqtls):1)),
    pvalue = -log10(cis_eqtls$pvalue[order(cis_eqtls$pvalue)]),
    gene = cis_eqtls$gene_name[order(cis_eqtls$pvalue)]
  )
df.ci$gene[20:nrow(df.ci)] <- NA

p <- ggplot(df.ci) +
  geom_point(aes(x = expected, y = pvalue), alpha = 0.5, size =  1) +
  ggrepel::geom_text_repel(aes(x = expected, y = pvalue, label = gene), 
                           size = 3, max.overlaps = 30)+
  geom_abline(intercept = 0, slope = 1, alpha = 0.5) +
  geom_line(data=df.ci,aes(expected, cupper), linetype = 2) +
  geom_line(data=df.ci,aes(expected, clower), linetype = 2) +
  cowplot::theme_cowplot() +
  ylab(substitute(paste("Observed -log10(", italic("P"), ")"))) +
  xlab(substitute(paste("Expected -log10(", italic("P"), ")"))) +
  ggtitle("Harbin", subtitle = paste0(nrow(cis_eqtls), " tested cis-eQTL"))
p
ggsave("figs/eQTL_AF_Harbin/QQ-plot_cis.eQTL_Harbin.png",
       p,
       width = 6, 
       height = 4)
```

# eQTLs genotype:condition interaction 

```{r}
cis_eqtls <- readRDS("results/Harbin_cis_eQTLs.sex.7PCs.RDS")
GE <- read.table("data/Harbin/MatrixEQTL/expr_g.txt", header = T)
meta.harbin <- readRDS("data/Harbin/RNAseq/meta_cov.7PCs.rds")
geno.o <- read.table("data/Harbin/MatrixEQTL/SNP.txt",
                     header = T,
                     row.names = 1, 
                     sep = "\t")

# Compute lm interactions for each SNP-gene pair
interactions.c <- lapply(1:nrow(cis_eqtls), function(i) {
  df <- meta.harbin
  snp <- as.character(cis_eqtls[i, "snps"])
  df$genotype <- t(geno.o[snp, ])
  df$genotype <- df$genotype[, 1]
  df$exprs <- GE[which(GE$id == cis_eqtls[i, "gene"]), -1] %>% 
    t() %>% 
    .[, 1]
  gene.name <- cis_eqtls$gene_name[i]

  # Create formula for linear model
  fs <- as.formula(paste("exprs ~ ", paste(
    c("genotype:condition", colnames(df)[-ncol(df)]), collapse = "+"
    )))
  m <- lm(formula = fs ,df)
  
  m2 <- summary(m)$coefficients[,4]
  m2["snp_gene.name"] <- paste0(snp,"_",gene.name)
  m2
}) %>% 
  do.call(rbind, .)

# make all columns numeric except SNP_gene
interactions.c.df <- as.data.frame(interactions.c)
interactions.c.df[, -ncol(interactions.c.df)] <- apply(
  interactions.c.df[, -ncol(interactions.c.df)], 2, as.numeric
  ) 

# FDR adjust
interactions.c.df$genotype.condition_FDR <- p.adjust(
  interactions.c.df$`genotype:condition`, method = "fdr"
  )
cis_eqtls <- cbind(cis_eqtls, interactions.c.df[, c("genotype:conditionSR", 
                                                    "genotype.condition_FDR")])
saveRDS(cis_eqtls, "results/Harbin_cis_interactions.7PCs.RDS")
```
